\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-06-30 tex/paper.tex
save

As a second example, we evaluate parameter inference for a generator producing
5-dimensional continuous data, as originally specified in Section 4.2. of
\citep{cranmer2015approximating}. More specifically, we consider the following
generative process:
\begin{itemize}
    \item $\mathbf{z} = (z_0, z_1, z_2, z_3, z_4)$, such that
        $z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
        $z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
        $z_2 \sim {\text{Mixture}}(\frac{1}{2}\,{\cal N}(\mu=-2, \sigma=1), \frac{1}{2}\,{\cal N}(\mu=2, \sigma=0.5))$,
        $z_3 \sim {\text{Exponential}(\lambda=3)}$, and
        $z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
    \item $\mathbf{x} = R  \mathbf{z}$, where $R$ is a fixed
    semi-positive definite $5 \times 5$ matrix defining a fixed projection
    of $\mathbf{z}$ into the observed space.
\end{itemize}
We consider observed data generated at the nominal values $\theta^* = (\alpha^*=1,\beta^*=-1)$.
The simulator parameters are modeled with a factored Gaussian
proposal distribution $q(\theta|\psi) = q(\alpha|\psi) q(\beta|\psi)$, where each component is
initialized with zero mean and unit variance. Hyper-parameters are set to $M=64$, $n_\text{critic}=100$, $\lambda=0.025$, $\gamma=10$ and
Adam configured with $\alpha=0.01$, $\beta_1=0.9$ and $\beta_2=0.999$.

@@

The top left plot in Figure~\ref{fig:weinberg} illustrates the resulting proposal
distributions $q(\theta|\psi)$ for $\gamma=0$ and $\gamma=5$ after adversarial
variational optimization. We see that the distributions both arrive in the
neighborhood of $\theta^*$, with a density highly more concentrated for
$\gamma=5$ than for $\gamma=0$.  Despite these differences and the relative
distance with $\theta^*$, both models closely match the observed data, as shown
in the top right plot of  Figure~\ref{fig:weinberg}, with again
slightly better results for the entropy penalized model. This suggests either a
relatively flat landscape around $\theta^*$ or that the observed data can in
this case also be reproduced with a smeared variation $p(\mathbf{x}|\psi)$ of
the simulator. Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that
for both $\gamma=0$ and $\gamma=5$ the variational upper bound $-U_d$ quickly
fall towards $0$, which indicates  convergence towards a distribution that the
critic cannot distinguish from the true distribution.

@@

>>>>

As a second example, we evaluate parameter inference for a generator producing
5-dimensional continuous data, as originally specified in Section 4.2. of
\citep{cranmer2015approximating}. More specifically, we consider the following
generative process:
\begin{itemize}
    \item $\mathbf{z} = (z_0, z_1, z_2, z_3, z_4)$, such that
        $z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
        $z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
        $z_2 \sim {\text{Mixture}}(\frac{1}{2}\,{\cal N}(\mu=-2, \sigma=1), \frac{1}{2}\,{\cal N}(\mu=2, \sigma=0.5))$,
        $z_3 \sim {\text{Exponential}(\lambda=3)}$, and
        $z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
    \item $\mathbf{x} = R  \mathbf{z}$, where $R$ is a fixed
    semi-positive definite $5 \times 5$ matrix defining a fixed projection
    of $\mathbf{z}$ into the observed space.
\end{itemize}
We consider observed data generated at the nominal values $\theta^* = (\alpha^*=1,\beta^*=-1)$.
The simulator parameters are modeled with a factored Gaussian
proposal distribution $q(\theta|\psi) = q(\alpha|\psi) q(\beta|\psi)$, where each component is
initialized with zero mean and unit variance.
Hyper-parameters are set to $M=64$, $n_\text{critic}=100$, $\lambda=0.025$, $\gamma=10$ and
Adam configured with $\alpha=0.01$, $\beta_1=0.9$ and $\beta_2=0.999$.
The network architecture of the critic is the same as in the previous example.

@@

The top left plot in Figure~\ref{fig:weinberg} illustrates the resulting
proposal distributions $q(\theta|\psi)$ for $\gamma=0$ and $\gamma=5$ after
adversarial variational optimization. We see that the distributions both arrive
in the neighborhood of $\theta^*$, with a density highly more concentrated for
$\gamma=5$ than for $\gamma=0$.  Despite these differences and the relative
distance with $\theta^*$, both models closely match the observed data, as shown
in the top right plot of  Figure~\ref{fig:weinberg}, with again slightly better
results for the entropy penalized model. This suggests either a relatively flat
landscape around $\theta^*$ or that the observed data can in this case also be
reproduced with a smeared variation $p(\mathbf{x}|\psi)$ of the simulator.
Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that for both
$\gamma=0$ and $\gamma=5$ the variational upper bound $-U_d$ quickly fall
towards $0$, which indicates  convergence towards a distribution that the critic
cannot distinguish from the true distribution. In summary, despite the intricate
relation betwen the observations $\mathbf{x}$, the latent variable $\mathbf{z}$
and the parameters $\theta$, this example demonstrates once again the efficacy
of adversarial variational optimization for parameter inference on black box
simulator.

@@

We would like to thank Lukas Heinrich for helpful comments regarding
the electron--positron annihilation simulation.
